### 2015-08-06 doc/manual/mscviewer.tex
Various fixes

%\usepackage[margin=1in]{geometry}
\usepackage{fixltx2e} 
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{tikz}
\usepackage{centernot}
\usepackage{listings}
\usetikzlibrary{shadows}
\usepackage[toc,page]{appendix}
\usepackage{etoolbox}
\usepackage{wrapfig}
\usepackage{tcolorbox}
\usepackage{lipsum}
\include{generated/titlepage}

@@

\section{Timestamps}
It is possible to associate timestamps to each event throught the \texttt{time}
key. The value can assume different formats, as shown in \texttt{lst4.msc},
visible in fig.~\ref{fig:gui-lst4}. \prog{} internally converts the units
indicated in the input files into nanoseconds. By default timestamps are shown
to the left of events in a \texttt{hh:mm:ss} format. The format can be changed
through the \texttt{view/Preferences...} menu item.

%\lstinputlisting[language=Java]{../../examples/lst4.msc}
\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.5\textwidth,natwidth=1024,natheight=600]{images/gui-ex4-time.png}
  \caption{Diagram with timestamps associated to events.}
  \label{fig:gui-lst4}
\end{figure}

@@

\section{Orphaned Interactions}
When collecting traces from a system it is possible that either the even
representing a cause or the one representing an effect for an interaction goes
missing. This can be caused by multiple reasons, such as the fact that a trace
has wrapped around (typically traces don't grow indefinitely, but rather wrap
around, loosing older information), or that traces for one of the two entity
involved in the interaction are not available. We will refer to interactions for
which only one of the two events is available as \defterm{Orphaned
Interactions}.
\prog{} visualizes orphaned interactions as dashed arrow stubs outgoing or
incoming into the only available event.

@@

\section{The Sequence Diagram area}
The sequence diagram area is situated on the top-right part of the GUI. Entities can be opened in this area by
double-clicking on them in the Entity Tree. Open entities can be rearranged by dragging them from their 
header.

\section{The Source View}
The source View is a tab in the bottom-right part of the GUI. This view shows the original log file that was 
parsed to create the sequence diagram model. Lines that contributed event information are highlighted in blue, 
while other lines are shown in black. Clicking on an event line selects the correspoding event, provided that the
owning entity was open in the sequence diagram area. Selecting an event in the sequence diagram causes the 
corresponding line in the Source View to be selected. Double-clicking on an event line in the Source View
causes the corresponding entity to be opened in the sequence diagram and the event selected.

\section{The Data View}
Events may have associated structured data. The data associated to the selected event is shown in the Data View,
which is located inside a tab in the bottomr-right part of the GUI.

\section{The Notes View}
In order to facilitate a debugging session the user can associate notes to events. An note associated with an
event is shown as a small yellow square on the event. Selecting the event and right-mouse-clicking the user can
checkmark the "Note Sticker" menu item to make the note visible as a sticker in the sequence diagram.
To create a note, the user selects the event and edit the note in the Notes View, which is situated in a tab
in the bottom-right part of the GUI.

is a tab in the bottom-right part of the GUI. This view shows the original log file that was 
parsed to create the sequence diagram model. Lines that contributed event information are highlighted in blue, 
while other lines are shown in black. Clicking on an event line selects the correspoding event, provided that the
owning entity was open in the sequence diagram area. Selecting an event in the sequence diagram causes the 
corresponding line in the Source View to be selected. Double-clicking on an event line in the Source View
causes the corresponding entity to be opened in the sequence diagram and the event selected.

@@

	\multicolumn{2}{l}{\textbf{Visualization Options}} \\
	\hline
	\tblimg{../../src/com/cisco/mscviewer/resources/icons/32x32/blocks.png}
	& \begin{minipage}[t]{0.8\columnwidth}
		Shows/hides blocks. 
	  \end{minipage}\\ \\
	\tblimg{../../src/com/cisco/mscviewer/resources/icons/32x32/time.png}
	& \begin{minipage}[t]{0.8\columnwidth}
		Shows/hides event timestamps
		\end{minipage}\\ \\
	\tblimg{../../src/com/cisco/mscviewer/resources/icons/32x32/label.png}
	& \begin{minipage}[t]{0.8\columnwidth}
		Shows/hides event labels
		\end{minipage}\\ \\
	\end{longtable}  
\end{center}

\chapter{Pyhton API}
One of the most powerful aspects of \prog{} is the integration with a Python interpreter
which allows to access programmatically both model as wel as GUI.

One use of this feature
consist in writing validation scripts that can browse model data and report whether a 
particular computational flow is present or not, it respects certain timing constraints, etc. 
Because the script is written once and then applied to traces produced by running the 
system multiple times, this scales the effort of an activity that otherwise would have to be
performed by one or more human beings.

@@

@@

The API is currently divided in the following modules:
\begin{itemize}
\item \texttt{msc.model} -- provides API to access entities, events and interactions in the model
\item \texttt{msc.gui} -- provides API to access the GUI 
\item \texttt{msc.utils} -- provides some utility API 
\item \texttt{msc.graph} -- provides API to access the graphing capabilities (experimental). 
\item \texttt{msc.flowdef} -- provides API to implement compact definition of expected flows and apply them
                              to the current model
\end{itemize}                              
\section{Model API}

@@

\section{GUI API}
\input{generated/python-api-gui.tex}

\section{Flows API}
\input{generated/python-api-flowdef.tex}

\section{Graph API}
\input{generated/python-api-graph.tex}

\chapter{Customizing \prog}
\section{Icons}
\section{Python Scripts}

\chapter{Flow Definition}
\section{Introduction}
\prog{} provides a powerful infrastructure supporting automated flow validation.
This means that a developer can write once a concise Python script describing an expected flow,
and apply such definition to models built from traces generated a runtime by a system. This
is the basis for flow-level verification automation, which is crucial to scale at least the first
level of triaging for a system.

\section{Your First Flow Definition}
The following is a simple flow-definition script:
\lstinputlisting{../../examples/flw1.py}

\appendix
\section{Useful Links}
\begin{tabular}{l l p{9.0cm}}

\end{tabular}

>>>>

%\usepackage[margin=1in]{geometry}
\usepackage{fixltx2e} 
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{tikz}
\usepackage{centernot}
\usepackage{listings}
\usetikzlibrary{shadows}
\usepackage[toc,page]{appendix}
\usepackage{etoolbox}
\usepackage{wrapfig}
\usepackage{tcolorbox}
\usepackage{lipsum}
\usepackage{dirtree}
\include{generated/titlepage}

@@

\section{Timestamps}
Timestamps can be associated to events throught the \texttt{time}
key. The value can assume different formats, as shown in \texttt{lst4.msc}:

\lstinputlisting[language=Java]{../../examples/lst4.msc}
.
Regardless of the unit specified in input, \prog{} converts internally the
timestamp into nanoseconds, and visualizes them in a format that can be chosen
by the user. The default format is \texttt{hh:mm:ss,ms}, but it can be 
changed through the the \texttt{Edit/Preferences...} menu item. Fig~\ref{fig:gui-lst4}
shows timestamps visualized near events.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.5\textwidth,natwidth=1024,natheight=600]{images/gui-ex4-time.png}
  \caption{Diagram with timestamps associated to events.}
  \label{fig:gui-lst4}
\end{figure}

@@

\section{Orphaned Interactions}
When collecting traces from a system it is possible that either the event
representing a cause or the one representing an effect for an interaction goes
missing. This can be caused by multiple reasons, such as the fact that a trace
has wrapped around (typically traces don't grow indefinitely, but rather wrap
around, loosing older information), or that traces for one of the two entity
involved in the interaction are not available. We will refer to interactions for
which only one of the two events is available as \defterm{Orphaned
Interactions}.
\prog{} visualizes orphaned interactions as dashed arrow stubs outgoing or
incoming into the only available event.

@@

\section{The Sequence Diagram}
The Sequence Diagram ais situated on the top-right part of the GUI. Entities can be opened in this area by
various user actions:
\begin{itemize}
\item double-clicking a node in the \textit{Entity Tree}, or pressing enter open the selected entity
\item double-clicking on a line in the \textit{Log File View}
\item programmatically, by Python scripts.
\end{itemize}

Open entities can be rearranged by dragging their \texttt{Header}, the rectangle containing the entity
name.

Width for an entity column can be increased or reduced by selecting the header and pressing repreatedly the
\keystroke{+} or \keystroke{-} keys.

\section{The Log File View}
The Log File View is a tab in the bottom-right part of the GUI. This view shows the original log file that was 
parsed to create the sequence diagram model. Lines that contributed event information are highlighted in blue, 
while other lines are shown in black. Clicking on an event line selects the correspoding event, provided that the
owning entity was open in the sequence diagram area. Selecting an event in the sequence diagram causes the 
corresponding line in the Source View to be selected. Double-clicking on an event line in the Source View
causes the corresponding entity to be opened in the sequence diagram and the event selected.

Clicking and dragging in the Log View causes text to be selected. When the mouse button is released the 
currently selected test is copied in the system clipboard.

\section{The Event Data View}
Events may have associated structured data. The data associated to the selected event is shown in the Data View,
which is located inside a tab in the bottomr-right part of the GUI.

\section{The Notes View}
In order to facilitate a debugging session the user can associate notes to events. A note associated with an
event is shown as a small yellow square on the event. Selecting the event and right-mouse-clicking the user can
checkmark the "Note Sticker" menu item to make the note visible as a sticker in the sequence diagram.
To create a note, the user selects the event and edit the note in the Notes View, which is situated in a tab
in the bottom-right part of the GUI.

@@

	\multicolumn{2}{l}{\textbf{Visualization Options}} \\
	\hline
	\tblimg{../../src/com/cisco/mscviewer/resources/icons/32x32/blocks.png}
	& \begin{minipage}[t]{0.8\columnwidth}
		Shows/hides blocks. 
	  \end{minipage}\\ \\
	\tblimg{../../src/com/cisco/mscviewer/resources/icons/32x32/time.png}
	& \begin{minipage}[t]{0.8\columnwidth}
		Shows/hides event timestamps
		\end{minipage}\\ \\
	\tblimg{../../src/com/cisco/mscviewer/resources/icons/32x32/label.png}
	& \begin{minipage}[t]{0.8\columnwidth}
		Shows/hides event labels
		\end{minipage}\\ \\
    \tblimg{../../src/com/cisco/mscviewer/resources/icons/32x32/note.png}
    & \begin{minipage}[t]{0.8\columnwidth}
        Shows/hides notes associated to events
        \end{minipage}\\ \\
	\end{longtable}  
\end{center}

\chapter{Customizing \prog}
\section{Preferences}
a number of parameters affecting the GUI can be modified in the \texttt{Preferences}
dialog accessible via the \texttt{Edit/Preferences...} menu item. These include:
\begin{itemize}
\item how to interpret event timestamps where units were not specified
\item the visualized format for timestamps in the sequence diagram
\item the format for entities headers in the sequence diagram
\item colors of various GUI elements   
\end{itemize}
When any of the parameters is modified the change is immediately visible in the GUI. If
the dialog is closed by pressing the \texttt{Ok} button, then those changes become permanent.
Pressing the \texttt{Cancel} button or closing the dialog via the OS-specific window-close button
results in the original parameters being restored.

The original parameters can also be restored without closing the dialog by pressing the \texttt{Reset}
button.

\section{Sessions}
While \texttt{Preferences} control program parameters irrespective of the loaded log files, some set 
of parameters can be associated to specific Log files loaded in the tool. These set of parameters, 
referrred to as \textit{Sessions}, are saved in XML files which can be reloaded later or even
transfer across different machines. Session information includes:
\begin{itemize}
\item Entities open in the sequence diagram
\item currently selected event or interaction
\item Notes associated to events
\item markers associated to events
\end{itemize}

\section{Passing Resource Sets to \prog{}}
At the time of execution \prog{} can be passed one or more directories, each one specifying a set of resources.
Resource directories are passed using the \texttt{-r} option. Directory paths should be separated using 
the OS-specific path separator character (\texttt{:} on Linux/OS-X and \texttt{;} on Windows). 
Each directory can contain one or both the following sub-directories:
\begin{itemize}
\item \texttt{renderer}: this directory can contain images in \texttt{.png} format. These images
are loaded by \prog and become available as renderer for events whose type matches the file names (without the extension).
\item \texttt{script}: this directory can contains Python scripts. These scripts are parsed by \prog and 
specially decorated functions become available for execution in the GUI. For more information 
on Python scripts see chapter~\ref{sec:python-api}. 
\end{itemize}

Consider the following example on Linux:
\begin{lstlisting}
mscviewer -r food;animals test.msc
\end{lstlisting}

\noindent where \texttt{test.msc} contains

\begin{lstlisting}
@event { "entity":"food",    "type":"bananas", "label":"bananas"}
@event { "entity":"food",    "type":"cake",    "label":"cake"}
@event { "entity":"animals", "type":"bee",     "label":"bee"}
@event { "entity":"animals", "type":"snail",   "label":"snail"}
\end{lstlisting}

\noindent and the current directory has the following directory structure:

\dirtree{% comment here is required as first char to be found must be a dot
.1 \texttt{.} . %
.2 animals/. % space after dot is required!
.3 renderer/. %
.4 bee.png. %
.4 snail.png. %
.2 food/. %
.3 renderer/. %
.4 bananas.png. %
.4 cake.png. %
}

This loads a sequence diagram where icons for various events are picked up from the \texttt{renderer}
subdirectories.

This example is present in the \texttt{examples/custom-icons} directory of the distribution, together with 
a \texttt{run.sh} script that runs it regardless of the OS in use.

\chapter{Python Integration}
One of the most powerful aspects of \prog{} is the integration with a Python interpreter
which allows to access programmatically both model as wel as GUI.

One use of this feature
consist in writing validation scripts that can browse model data and report whether a 
particular computational flow is present or not, it respects certain timing constraints, etc. 
Because the script is written once and then applied to traces produced by running the 
system multiple times, this scales the effort of an activity that otherwise would have to be
performed by one or more human beings. A tutorial on how to create flow validation scripts
using the APIs described in this chapter is provided in chapter~\ref{chp:flowdef}

@@

The interaction between the user Python script and the tool is done through APIs 
collected in a few Python files the script can import, and described in details 
in \ref{sec:python-api}.

\section{\textit{Hello World} of Python Integration}

Let's start with the smallest possible \textit{Hello World} program in Python (\texttt{examples/hello.py}):

\lstinputlisting{../../examples/hello.py}

This can be executed by \prog{} as follows:

\begin{verbatim}
$ mscviewer -b examples/hello.py examples/lst1.msc 
 Hello World
\end{verbatim}

Here we're invoking \prog{} in batch mode, passing the \texttt{hello.py} script and
a log file. The log file is not even used in this case, and the script is simply
passed to the embedded Python interpreter which runs it, producing the output on
the console. Not particularly exciting.

Let's examine next how we can interact with the model created from the log file, as
demonstrated by \texttt{examples/info.py}:

\lstinputlisting{../../examples/info.py}

\begin{verbatim}
$ bin/mscviewer -b examples/info.py examples/lst1.msc
 entity= producer label= start
 entity= producer label= produce
 entity= consumer label= consume
\end{verbatim}

This example uses the Python API to interact with the model created from the log file, 
printing out for each event the entity path and the label. The extensive set of APIs 
provided to interact with the model as well as the GUI is documented in section
\ref{sec:python-api}.

In addition to executing a single Python program in batch mode, it is possible
to define any number of Python functions that can be explicitly invoked by the
user within the GUI. To demonstrate this, let's consider the \texttt{examples/hello1/hello.py}
file:

\lstinputlisting{../../examples/hello1/hello.py}

In order for functions to be visible by the gui we need to place the Python file declaring them
into a package, which we can achieve by creating an empty \texttt{\_\_init\_\_.py} file in the 
directory containing the file. For this reason, the \texttt{examples/hello1} directory has
the following structure:

\hfill \break
\dirtree{% comment here is required as first char to be found must be a dot
.1 \texttt{hello1}. %
.2 \texttt{\_\_init\_\_.py}. % space after dot is required!
.2 \texttt{hello.py}. % space after dot is required!
}
\hfill \break

We can now run \prog{} as follows:

\begin{verbatim}
mscviewer -p examples/hello1 examples/lst1.msc
\end{verbatim}

Once the GUI comes up, click on the \texttt{scripts} tab. The tool will parse the Python
modules available in the path, and the visualized tree should contain a \texttt{hello1}
node representing the package. Expanding such node should reveal a leaf node representing
the \texttt{hello\_world()} function. Notice that in order for a function to show up in the
GUI tree the function has to be decorated by the \texttt{@msc\_fun} decorator. This 
allow to have various utility functions in the file that do not show up as first class
citizens in the GUI.

In order to execute the function, just double-click on its node in the tree. In this 
particular case the function takes no arguments, so it will be immediately executed producing
the expected \texttt{Hello World} output on the console.

Let's now consider \texttt{examples/hello2/hello.py}:

\lstinputlisting{../../examples/hello2/hello.py}

This example provides three functions that will be listed in the GUI:
\begin{itemize}
\item \texttt{hello\_something()} expects an argument of type string and prints the concatenation of \texttt{'Hello '}
followed by such string. The first time we double-click on this node a dialog will open allowing us to 
provide a value for the argument. We can type \texttt{'World'} (including the apices) and press \texttt{OK}. 
This time the output will not be produced on the console, but rather in the \texttt{Script Console} tab in the 
bottom-right are of the GUI. Double-clicking on the same node again will not open the dialog again, as \prog{}
remembers the previosly passed value and uses it again.  In order to open the dialog again and change the 
value keep the \keystroke{SHIFT} key pressed while double-clicking on the node. 
\item \texttt{hello\_world()} calls the previous function passing the value of \texttt{'World'} for the argument.
\item \texttt{hello\_world\_or\_not()} has a keyword parameter. Keyword parameters have a default
value, hence double-clicking on the node executes the function with such default value for the
parameter. Again, to change such value keep the \keystroke{SHIFT} key pressed while double-clicking.
\end{itemize}

It should be noted that this particular example follows a common pattern which allows a script
to be executed in batch mode, or to provide functions to be called in interactive mode. The pattern
consists in the presence of an \texttt{if \_\_name\_\_ == '\_\_main\_\_'} statement which 
is executed when in batch mode, calling a function with default values. Also, the pattern
makes use of the \texttt{msc\_print} function to produce output: this function will direct
its output to the GUI \texttt{Script Console} view when in interactive mode, while 
falling back to printing on the console when in batch mode. We can run the 
script in batch mode with

\begin{verbatim}
mscviewer -b examples/hello2/hello.py examples/lst1.msc
\end{verbatim}

\

\section{Pyhton API}
\label{sec:python-api}

@@

The API is currently divided in the following modules:
\begin{itemize}
\item \texttt{msc.model} -- provides API to access entities, events and interactions in the model
\item \texttt{msc.gui} -- provides API to access the GUI 
\item \texttt{msc.utils} -- provides some utility API 
\item \texttt{msc.graph} -- provides API to access the graphing capabilities (experimental). 
\item \texttt{msc.flowdef} -- provides API to implement compact definition of expected flows and apply them
                              to the current model
\end{itemize}

\subsection{Model API}

@@

\subsection{GUI API}
\input{generated/python-api-gui.tex}

\subsection{Flows API}
\label{sec:flow-api}
\input{generated/python-api-flowdef.tex}

\subsection{Graph API}
\input{generated/python-api-graph.tex}

\section{Flow Definition}
\label{chp:flowdef}

\prog{} provides a powerful infrastructure supporting automated flow validation.
This means that a developer can write once a concise Python script describing an expected flow,
and apply such definition to models built from traces generated a runtime by a system. This
is the basis for flow-level verification automation, which is crucial to scale at least the first
level of triaging for a system.

\subsection{Your First Flow Definition}
File \texttt{examples/flw1.py} contains a simple flow definition script. Let's look at it in detail:
\lstinputlisting{../../examples/flw1.py}

\noindent Lines 1, 2 import some packages required for flow definition. These are defined in Python files
contained in the \texttt{resources/default} directory of the \prog{} install directory.

Lines 5-7 define a function returning a flow definition. The flow definition is specified as a sequence (\texttt{fseq})
of two events (\texttt{fev}). The two parameters for each \texttt{fev()} element are respectively the entity id and
the label of the events. A model containing an ordered sequence of two events on an entity named \textit{foo}, the 
first one with label \texttt{event\_a} and the second with label \texttt{event\_b} does contain the flow specified
by function \texttt{get\_flow()}. Function \texttt{get\_flow()} returns the defined flow as an object.

Function
\texttt{find\_flow()} at line 12 calls such function to retrieve the flow definition and repeatedly calls function
\texttt{match()} on it. the \texttt{match()} function attempts to match the flow definition to the model
currently loaded in \prog{}, starting with the model event with the index specified as the \texttt{start\_event\_idx}
parameter. In case of match, the function returns the index of the last event in the model matching one of the elements
in the flow definition, plus one. If the entire flow definition is not found in the model, then the function throws a
FlowError. If passed a starting event index corresponding to the total number of events the function returns -1.

The \texttt{examples/flw1-match.msc} is an example of a log file containing the specified flow:
\lstinputlisting{../../examples/flw1-match.msc}

However, note that the \texttt{examples/flw1-match.msc} file also contains the specified flow:

\lstinputlisting{../../examples/flw1-match2.msc}

This is the case because \texttt{fseq} requires events to be in the specified order, but
not to be contiguous in the file, or even within the entity.

We can test the scripts by running \prog{} in batch mode:

\begin{verbatim}
$ mscviewer -b examples/flw1.py examples/flw1-match.msc 
 Found match!
$ mscviewer -b examples/flw1.py examples/flw2-match.msc 
 Found match!
$ bin/mscviewer -b examples/flw1.py examples/flw1-nomatch.msc
 No match found ( element (foo, event_a) not found in model. )
$\end{verbatim}

\subsection{Model Elements for a Flow}

In addition to knowing that a certain flow is present in a model, you may want to discover which 
\texttt{fev} elements match which mdodel events. This can be achieved using the \texttt{get\_model\_map()}
function, as illustrated by the \texttt{examples/flw2.py} example:

\lstinputlisting{../../examples/flw2.py}

The \texttt{get\_model\_map()} function returns a map where each \texttt{fev} in the flow definition is a key,
and the event matching it is the value associated to the key. Various pieces of information 
can be stracted from the event using the \texttt{event\_} API defined in the \texttt{msc.model} module.
In this example, we print the line number where the event is defined in the input file:

\begin{verbatim}
$ bin/mscviewer -b examples/flw2.py examples/flw1-match.msc
 found match!
 fev(foo, event_b) : 2
 fev(foo, event_a) : 1
$ bin/mscviewer -b examples/flw2.py examples/flw1-match2.msc
 found match!
 fev(foo, event_b) : 4
 fev(foo, event_a) : 1
\end{verbatim}

\appendix
